\singlespacing \vspace{-1.6cm} \LARGE Covariation among reproductive traits in flowering plants shapes interactions with floral visitors

### 2022-10-04 Manuscript/Title_Page.Rmd
edits

```{=tex}
\doublespacing
\normalsize
```

# Abstract

Globally, plants display enormous variation in life history strategies and trait combinations. However, evidence suggests that evolutionary and physiological constraints limit the number of plant ecological strategies. Although there have been recent advances in understanding correlations among plant traits, reproductive traits are rarely considered, despite their key role in shaping plant life-history strategies and interactions with floral visitors. Here, using a global dataset of 16 reproductive traits for 1,506 species, we investigate the reproductive spectrum of flowering plants to identify how it shapes interactions with floral visitors. We show that over 50% of all trait variation is explained by the first two reproductive axes, which represent the negative correlation between flower number and flower size, and the negative correlation between autonomous selfing and floral display size. In addition, these reproductive axes were associated with different numbers of interacting partners and visitation rates of the distinct floral visitor guilds. However, the relatively small amount of variance explained by the different reproductive axes highlight the need to incorporate other factors along with reproductive traits to fully explain large scale patterns of plant-pollinator interactions. Our study identifies the major reproductive trait correlations in flowering plants and their role in shaping networks of plant-pollinator interactions at a global scale.

\textbf{Keywords:} life-history strategies \| plant reproductive traits \| plant-pollinator interactions

```{=tex}
\doublespacing
\vspace{5mm}
\normalsize
```
```{r setup, include=FALSE, cache=FALSE, message = FALSE}

library("knitr")

### Chunk options: see http://yihui.name/knitr/options/ ###
## Text results
opts_chunk$set(echo = FALSE, warning = TRUE, message = TRUE)

## Code decoration
opts_chunk$set(tidy = TRUE, comment = NA, highlight = TRUE)

## Cache
# opts_chunk$set(cache = 2, cache.path = "output/cache/")

## Plots
opts_chunk$set(fig.path = "output/figures/")

```

```{r knitcitations, cache = FALSE}
library(knitcitations)
cleanbib()   
cite_options(citation_format = "pandoc")

```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

# 1. Introduction

Flowering plants display an astonishing diversity of reproductive structures [@barrett2002; @schiestl2013] that shape their interactions with animal pollinators [@fenster2004; @dellinger2020]. The variety of reproductive strategies used by different plant species, from specialised reliance on a single pollinator, to full self-pollination, is incredibly rich. However, not all reproductive trait combinations in plants are possible due to evolutionary and ecological constraints [@stearns1989; @agrawal2020]. Despite recent advances in theoretical and empirical understanding of correlations between plant reproductive traits at large ecological scales [@salguero2016; @friedman2020; @paterno2020; @roddy2021; @song2022], studies typically focus on just a few traits, often in isolation. Thus, to develop a comprehensive understanding of the plant reproductive spectrum of trait variation, as done recently for root [@laughlin2021], leaf [@wright2004] and wood [@chave2009] related traits, we require a multi-trait perspective with broad geographical coverage. Importantly, characterization of the plant reproductive spectrum could reveal key reproductive strategies among different species [@agrawal2020] and improve knowledge of associations between specific reproductive traits and pollinators [@roddy2021].

With the recent availability of large trait databases [e.g., TRY @kattge2011;  COMPADRE @salguero2015], there has been increased research on plant ecological strategies, which has produced identification of global patterns in plant form and function [@salguero2016; @diaz2016; @bruelheide2018; @carmona2021]. However, studies that look at multiple traits tend to overlook plant reproductive traits [@evojtko2020; @ruger2018; @roddy2021], and focus on trait correlations in other plant organs such as morphological and physiological leaf traits [@wright2004; @shipley2006; @donovan2011; @osnas2013]. Despite the lack of a holistic understanding that depicts reproductive trait covariation patterns, there are widely recognized associations between plant reproductive trait pairs such as the negative correlation between flower size and flower number [@sargent2007; @kettle2011; @vojtko2022]; the positive association between flower size and outcrossing rate [@goodwillie2010]; or the positive correlation between outcrossing rate and lifespan [@barrett2003; @munoz2016; @moeller2017]). Although identification of these trait correlations has allowed progress toward a conceptual framework that integrates different floral trait relationships [@roddy2021], we still lack empirical understanding of how these associations among floral and other reproductive traits (e.g., reproductive biology) shape plant life history strategies.

Trait-based approaches [@fenster2004; @rosas2014] and trait-matching analyses [@stang2009; @bartomeus2016] are powerful for exploring the underlying mechanisms driving plant-pollinator interactions. For example, plant traits can define species' network roles [e.g., specialists vs generalists @tur2013] and plant species that occupy reproductive trait space extremes are likely to be more specialised and display tighter trait-matching with pollinators [@junker2013; @coux2016]. Indeed, the degree of morphological matching between plants and floral visitors often determines whether or not interactions occur, and consequently influences  the broader network structure [@stang2009; @ibanez2012]. However, it is still unclear how specific reproductive traits (e.g., mating or compatibility system) influence plant-pollinator associations [@sargent2008; @tur2013; @devaux2014]. Indeed, both morphology and mating system can determine a species' functional role within pollination networks, and the combination of several traits increases power for predicting network structure [@eklof2013]. Because the species position within the multidimensional trait space can result in different interaction patterns or adaptations to pollinators [see @dehling2016 for a purely morphological approach], the use of the reproductive trait space to explore plant-pollinator interactions could help to identify species functional roles and associations of traits with specific pollinators (i.e., pollination syndromes).

Here, we explore the spectrum of reproductive traits in flowering plants, within a global set of plant-pollinator networks, to identify reproductive trait covariation and how this influences the network of interactions between plants and floral visitors. First, we identify the major axes of reproductive trait variation and correlations among traits for plant species. Then, we investigate the associations between plant species’ positions in reproductive trait space and the strength of their interactions with different floral visitor guilds within networks. Finally, we determine how the main axes of reproductive trait variation influence plant species' functional roles within pollination networks, using a range of complementary interaction network metrics.

# 2. Methods

## (a) Plant-pollinator network studies

We selected 28 studies from 18 different countries that constituted a total of 64 plant-pollinator networks (see table S1 and figure S1). These studies recorded plant-pollinator interactions in natural systems and were selected so that we had broad geographical representation across different ecological communities. Although these studies differ in sampling effort and methodology, all studies provided information about plant-pollinator interactions (weighted and non-weighted), which we used to build a database of plant species that are likely to benefit, to some degree, from animal pollination. Many of these networks are freely available either as published studies [e.g., @carvalheiro2014; @olesen2007; @fortuna2010] or in online archives [e.g., Mangal @poisot2016;  The Web of Life @fortuna2010]. In total, our network dataset constituted 60 weighted (number of visits) and four unweighted (interaction presence-absence) networks, each sampled at a unique location and year, as well as eight meta-networks where interactions were pooled across several locations and multiple years.

## (b) Taxonomy of plants and pollinators

All species names, genera, families and orders were retrieved and standardised from the taxonomy data sources NCBI (<https://www.ncbi.nlm.nih.gov/taxonomy>) for plants and ITIS (<https://www.itis.gov/>) for pollinators, using the R package *taxize* [@chamberlain2020]. We filled the 'not found' searches manually using <http://www.theplantlist.org/> and <http://www.mobot.org/> for plants and <http://www.catalogueoflife.org/> for floral visitors.

## (c) Plant traits

We selected a total of 19 quantitative and categorical functional traits that included reproductive traits and traits related with plant form and size (see table 1). Reproductive traits consisted of those related to floral and reproductive biology: (i) floral traits included those related to the size of floral organs (e.g., style length), floral display size (e.g., flower number) and floral rewards (e.g., pollen quantity), while (ii) reproductive biology traits represented the plant reproductive system (i.e., the breeding, mating and compatibility system). The remaining traits were related to plant size and form (i.e., plant height, lifespan and life form), which are commonly used to characterise the fast-slow continuum of plant trait variation [@salguero2016]. For each plant species, we undertook an extensive literature and online search for all traits across a wide range of resources (plant databases, online floras, books, journals and images). From a matrix comprising a total of 28,614 possible cells (considering all traits and plant species: 19 columns × 1,506 species), we were able to fill 23,354 cells (81.6% of the dataset, see figure S2 for missing value information for each trait). An extended description of each trait and how it was obtained can be found in Appendix S1.

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.height=1, fig.width=14}
setwd("~/R_Projects/Reproductive Traits")

library(kableExtra)

library(dplyr)
library(readxl) #read excel file (trait data)
library(dplyr) #data manipulation
library(visdat) #VISUALIZE MISSING DATA
library(naniar)
library(tidyverse)
########################################################################################################################################################
#1) READ TRAIT DATA
########################################################################################################################################################
#load data
trait_data <- read_excel("Data/Trait_data_raw/Trait_data_final.xlsx",na = "NA")
########################################################################################################################################################
#2) A) CLEAN DATA AND B) RENAME LEVELS FOR IMPUTATION 
########################################################################################################################################################
#####
#A) CLEAN DATA
#####
#select just filled rows
trait_data_1 <- trait_data[1:1712,] #It may be a more elegant way but this does the job, 1712 rows of data

#filter data, select species with flower level info and capitulum
trait_filtered <- filter(trait_data_1, Info_level == "flower" |  Info_level == "capitulum")
#remove NA's and duplicated species|Some spp are repeated once the spp names are standardize with taxize
trait_filtered$Species_all[trait_filtered$Species_all=="NA"]<-NA 
trait_filtered_1 <- trait_filtered[!is.na(trait_filtered$Species_all),]
trait_filtered_2 <- trait_filtered_1[!duplicated(trait_filtered_1$Species_all),]

#Select columns to work with
t <- trait_filtered_2[c("Species_geonet","Order_all","Family_all","Genus_all","Species_all","Breeding_system","Compatibility_system",
                        "Autonomous_selfing_level","Autonomous_selfing_level_fruit_set", "Flower_morphology", "Flower_symmetry", 
                        "Flowers_per_plant","Floral_unit_width", "Corolla_diameter_mean", "Corolla_length_mean","Style_length", "Ovule_number", 
                        "life_form", "lifespan","Plant_height_mean_m","Nectar_presence_absence","Nectar_ul","Nectar_mg","Nectar_concentration","Pollen_per_flower")]

########################################################################################################################################################
#####
#B) RENAME LEVELS FOR IMPUTATION
#####
#Breeding_system| For the purpose of this study we are not going to consider other complex breeding systems and just focus on 
#Hermaphroditism, dioecy and monoecy, the closest levels are grouped within each category
t$Breeding_system <- as.character(t$Breeding_system)
t$Breeding_system[t$Breeding_system=="androdioecious"] <- "dioecious"
t$Breeding_system[t$Breeding_system=="androgynodioecious"] <- "dioecious"
t$Breeding_system[t$Breeding_system=="andromonoecious"] <- "monoecious"
t$Breeding_system[t$Breeding_system=="gynodioecious"] <- "dioecious"
t$Breeding_system[t$Breeding_system=="gynoecious"] <- "monoecious"
t$Breeding_system[t$Breeding_system=="gynomonoecious"] <- "monoecious"
t$Breeding_system[t$Breeding_system=="gynomonoecious"] <- "monoecious"
t$Breeding_system[t$Breeding_system=="monoecious_dioecious"] <- "monoecious"
t$Breeding_system[t$Breeding_system=="polygamo-dioecious"] <- "dioecious"
t$Breeding_system[t$Breeding_system=="protandrous"] <- "hermaphrodite"
t$Breeding_system[t$Breeding_system=="protogynous"] <- "hermaphrodite"
t$Breeding_system[t$Breeding_system=="subdioecious"] <- "dioecious"
t$Breeding_system[t$Breeding_system=="submonoecious"] <- "monoecious"
#Change to capital letters| There are already levels with lower case
t$Breeding_system[t$Breeding_system=="monoecious"] <- "Monoecious"
t$Breeding_system[t$Breeding_system=="dioecious"] <- "Dioecious"
t$Breeding_system[t$Breeding_system=="hermaphrodite"] <- "Hermaphrodite"
t$Breeding_system <- as.factor(t$Breeding_system)

#Now 3 levels
#Flower_morphology
t$Flower_morphology <- as.character(t$Flower_morphology)
t$Flower_morphology[t$Flower_morphology=="bowl"] <- "open"
t$Flower_morphology[t$Flower_morphology=="dish"] <- "open"
t$Flower_morphology[t$Flower_morphology=="exposed"] <- "open"
t$Flower_morphology[t$Flower_morphology=="spadix"] <- "spike"
t$Flower_morphology[t$Flower_morphology=="open"] <- "Open"
t$Flower_morphology[t$Flower_morphology=="brush"] <- "Brush"
t$Flower_morphology[t$Flower_morphology=="campanulate"] <- "Campanulate"
t$Flower_morphology[t$Flower_morphology=="capitulum"] <- "Capitulum"
t$Flower_morphology[t$Flower_morphology=="funnelform"] <- "Funnelform"
t$Flower_morphology[t$Flower_morphology=="papilionaceous"] <- "Papilionaceous"
t$Flower_morphology[t$Flower_morphology=="spike"] <- "Spike"
t$Flower_morphology[t$Flower_morphology=="tube"] <- "Tube"
t$Flower_morphology <- as.factor(t$Flower_morphology)

#Life span
t$lifespan <- as.character(t$lifespan )
t$lifespan[t$lifespan=="annual"] <- "short_lived"
t$lifespan[t$lifespan=="biennial"] <- "short_lived"
t$lifespan[t$lifespan=="short_lived"] <- "Short lived"
t$lifespan[t$lifespan=="perennial"] <- "Perennial"
t$lifespan <- as.factor(t$lifespan )

########################################################################################################################################################
#3) EXPLORE PATTERNS OF MISSING DATA
########################################################################################################################################################

#We convert cualitative data to quantitative to reduce missing data on this column and hence conduct the imputation 
#use ifelse base r does not work with these NA'S Even with work around. base r does not like the na option of read excel
t$Autonomous_selfing_level_fruit_set <- ifelse(t$Autonomous_selfing_level %in% c("high") & is.na(t$Autonomous_selfing_level_fruit_set), 88, t$Autonomous_selfing_level_fruit_set)
t$Autonomous_selfing_level_fruit_set <- ifelse(t$Autonomous_selfing_level %in% c("medium") & is.na(t$Autonomous_selfing_level_fruit_set), 50.5, t$Autonomous_selfing_level_fruit_set)
t$Autonomous_selfing_level_fruit_set <- ifelse(t$Autonomous_selfing_level %in% c("low") & is.na(t$Autonomous_selfing_level_fruit_set), 13, t$Autonomous_selfing_level_fruit_set)
t$Autonomous_selfing_level_fruit_set <- ifelse(t$Autonomous_selfing_level %in% c("none") & is.na(t$Autonomous_selfing_level_fruit_set), 0, t$Autonomous_selfing_level_fruit_set)

#t$Nectar_ul adding it here just to see how many cells 
#Now we just select presence/absence of nectar
#select columns of interest
t <- t[c("Species_geonet","Order_all","Family_all","Genus_all","Species_all","Breeding_system","Compatibility_system","Autonomous_selfing_level",
         "Autonomous_selfing_level_fruit_set", "Flower_morphology", "Flower_symmetry", "Flowers_per_plant", "Floral_unit_width",
         "Corolla_diameter_mean", "Corolla_length_mean", "Style_length", "Ovule_number", "life_form", "lifespan",
         "Plant_height_mean_m","Nectar_presence_absence","Nectar_ul","Nectar_mg","Nectar_concentration","Pollen_per_flower")]

########################################################################################################################################################
#4) CALCULATE PHYLOGENETIC DISTANCE TO CORRECT WITH EIGENVALUES IN THE IMPUTATION
########################################################################################################################################################
#Convert to factor 
cols <- c("Order_all", "Family_all", "Genus_all", "Species_all", "Compatibility_system", "Autonomous_selfing_level", "Flower_morphology",
          "Flower_symmetry", "life_form", "lifespan","Nectar_presence_absence")
t[cols] <- lapply(t[cols], factor)  ## as.factor() could also be used

cols.num <- c("Family_all","Genus_all","Species_all")
t[cols.num] <- sapply(t[cols.num],as.character)
t$Species_all <- gsub("Species_all_", "", t$Species_all)

t <- t[!t$Species_all == "Pinus luchuensis", ]   # remove gymnosperm species

#remove species that are not until species level
#ALL THE SPECIES WITH SP. ARE DELETD
t$Species_geonet <- gsub("M_PL_.*","",t$Species_geonet) #subsitute partial string for nothing
t$Species_geonet <- gsub(" $","", t$Species_geonet, perl=T) #remove trailing spaces
t$Species_geonet <-  sub('sp.1$', 'sp.', t$Species_geonet)#PREPARE ALL SP TO SP.
t$Species_geonet <- sub('sp.2$', 'sp.', t$Species_geonet)#PREPARE ALL SP TO SP.
t$Species_geonet <- sub('sp$', 'sp.', t$Species_geonet) #PREPARE ALL SP TO SP.
t$Species_geonet <- sub("sp.$", "DELETE", t$Species_geonet) #change all sp. for DELETE
t <- t[- grep("DELETE",  t$Species_geonet),] #remove species with "DELETE"
#CHECK LEVELS

#Make these NA's as NA
t$Species_all[t$Species_all=="NA"] <- NA
t$Family_all[t$Family_all=="NA"] <- NA
t$Genus_all[t$Genus_all=="NA"] <- NA
#remove NA's
t <- t[!is.na(t$Family_all),]
t <- t[!is.na(t$Species_all),]
t <- t[!is.na(t$Genus_all),]

missing.values <- t %>%
  gather(key = "key", value = "val") %>%
  mutate(isna = is.na(val)) %>%
  group_by(key) %>%
  mutate(total = n()) %>%
  group_by(key, total, isna) %>%
  summarise(num.isna = n()) %>%
  mutate(pct = num.isna / total * 100)

n_of_records <- subset(missing.values, isna=="FALSE")

Trait <- c("Plant height", "Flower width", "Flower length", "Inflorescence width", "Style length", "Ovules per flower",
           "Flowers per plant", "Microliters of nectar", "Milligrams of nectar", "Nectar concentration", "Pollen grains per flower", "Autonomous selfing (fruit set)")

Trait1 <- linebreak(c("Lifepan", "Life form", "Flower shape", "Flower symmetry", "Nectar", "Compatibility system", "Breeding system"),align = c("l"))

#T.cat <- linebreak(c("Short-lived \n Perennial", "Herb \n Shrub #\n Tree", "Brush \n Campanulate \n Capitulum \n Open \n #Papilionaceous \n Tube",  "Actinomorphic \n Zygomorphic",
#           "Presence \n Absence", "Self-incomp. \n Part. #self-comp. \n Self-comp.", "Hermaphrodite \n Monoecious \n #Dioecious"),align = c("l"))
#
T.cat <- linebreak(c("Short-lived and perennial", "Herb, shrub and tree", "Brush, campanulate, capitulum,\n open, papilionaceous and tube",  "Actinomorphic and zygomorphic",
           "Presence and absence", "Self-incompatible, partially self-compabtile \n and self-compatible", "Hermaphrodite, monoecious and dioecious"), align = c("l"))

Number_of_records_quantitative <- c(n_of_records$num.isna[n_of_records$key=="Plant_height_mean_m"], n_of_records$num.isna[n_of_records$key=="Corolla_diameter_mean"],
  n_of_records$num.isna[n_of_records$key=="Corolla_length_mean"], n_of_records$num.isna[n_of_records$key=="Floral_unit_width"],
  n_of_records$num.isna[n_of_records$key=="Style_length"], n_of_records$num.isna[n_of_records$key=="Ovule_number"],
  n_of_records$num.isna[n_of_records$key=="Flowers_per_plant"], n_of_records$num.isna[n_of_records$key=="Nectar_ul"],
  n_of_records$num.isna[n_of_records$key=="Nectar_mg"], n_of_records$num.isna[n_of_records$key=="Nectar_concentration"], n_of_records$num.isna[n_of_records$key=="Pollen_per_flower"],
  n_of_records$num.isna[n_of_records$key=="Autonomous_selfing_level_fruit_set"])

Number_of_records_categorical <- c(n_of_records$num.isna[n_of_records$key=="lifespan"], n_of_records$num.isna[n_of_records$key=="life_form"],
                                   n_of_records$num.isna[n_of_records$key=="Flower_morphology"], n_of_records$num.isna[n_of_records$key=="Flower_symmetry"],
                                   n_of_records$num.isna[n_of_records$key=="Nectar_presence_absence"], n_of_records$num.isna[n_of_records$key=="Autonomous_selfing_level"],
                                   n_of_records$num.isna[n_of_records$key=="Compatibility_system"], n_of_records$num.isna[n_of_records$key=="Breeding_system"],
                                   rep("",4))

T.range <- c("0.01-30m", "0.50-205mm", "0.20-195mm", "0.80-300mm", "0-138mm", "1-25000", "1-1.8 x $10^5$", "0-160$\\mu$l", "0-6.1mg", "0-100$\\%$", "13-2 x $10^8$", "0-100$\\%$")

dat_1 <- data.frame(Trait,T.range)
dat_2 <- data.frame(Trait1, T.cat)

colnames(dat_1) <- c("Quantitative traits", "Numerical range")
colnames(dat_2) <- c("Categorical traits", "Categories")

#kable(list(dat_1, dat_2), booktabs = T, linesep = #"\\addlinespace", escape = FALSE,caption = "\\textbf{Table 1 | #Quantitative and categorical traits used in this study.}") %>%
#  kable_styling(full_width = F,font_size = 10) %>%  #column_spec(1:4, 
#              width = "3in", 
#              latex_valign = "m")
#
#

colnames(dat_1) <- c("Traits", "Numerical range/Categories")
colnames(dat_2) <- c("Traits", "Numerical range/Categories")

d <- rbind(dat_1,dat_2)

kable(d, booktabs = T, linesep = "\\addlinespace", escape = FALSE,caption = "\\textbf{Table 1.} Quantitative and categorical traits used in this study.") %>%
  kable_styling(full_width = F,font_size = 12) %>% row_spec(0, bold=T) %>% pack_rows("Quantitative", 1, 12) %>%
pack_rows("Categorical", 13, 19)

```

\doublespacing

## (d) Phylogenetic Distance

We calculated the phylogenetic distance between all plant species considered using the function *get_tree* from the package *rtrees* (<https://github.com/daijiang/rtrees>), which downloads phylogenetic distances from the extended R implementation of the Open Tree of Life [@smith2018; @jin2019].

## (e) Data Imputation

To avoid excluding species with trait missing values, we imputed trait missing values with the help of the function *missForest* [@stekhoven2012] which allows imputation of datasets with continuous and categorical variables. We accounted for the phylogenetic distance among species on the imputation process by including the eigenvectors from principal component analysis of the phylogenetic distance (PCoA), which has been shown to improve the performance of *missForest* [@penone2014]. To extract the eigenvectors, we used the function *PVRdecomp* from the package *PVR* [@santos2018] based on a previous conceptual framework that considers phylogenetic eigenvectors [@diniz-filho2012]. We conducted two different imputations, one for the full set of species (1,506 species: 5.79% of missing values), and excluded nectar and pollen traits because of the high percentage of missing values (figure S2), and a second one for the subset of species with data for pollen per flower and microliters of nectar (755 species: 8.01% of missing values). To corroborate that our imputation of missing values did not affect our results, we evaluated the reproductive spectrum (see section below) with and without imputed values and results for the two datasets were consistent (Fig S3 and Fig S4).

## (f) Plant strategies analysis

We explored the association between the different quantitative plant traits using a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width, respectively (Pearson's correlation = 0.72, *P* \< 0.01 and Pearson's correlation = 0.36, *P* \< 0.01), and thus we avoided overemphasise flower size on the spectrum of trait variation. Prior to the analyses, we excluded outliers and standardised the data. Specifically, due to the high sensitivity of dimensionality reduction to outliers [@serneels2008; @legendre2012], we excluded values outside the 2.5th--97.5th percentile range, and thus our final dataset had 1,236 species. Then, we log transformed all continuous variables to reduce the influence of extreme values and then z-transformed (X= 0, SD=1) these variables so that they were within the same numerical range, as recommended for principal component analysis [@legendre2012]. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the statistical association of the different qualitative traits with the main axes of trait variation using an ANOVA and Tukey test. We performed the pPCA using the function *phyl.pca* from the package *phytools* [@revell2012] with the method lambda ($\lambda$) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode “covariance” because values for each variables were on the same scale following transformation [@abdi2010]. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity data. For this, we included all quantitative traits included in the main pPCA, plus pollen grains and microlitres of nectar per flower.

## (g) Phylogenetic signal

We calculated the phylogenetic signal of the different quantitative traits on the imputed dataset with the full set of species (N = 1,506) with the package *phytools* [@revell2012] and used Pagel's $\lambda$ as a measurement of the phylogenetic signal. However, for pollen and nectar traits, phylogenetic signal was calculated only on the subset of species that had quantitative information for these traits (N = 755).

## (h) Network analyses

### (i) Visitation patterns

We used Bayesian modelling (see details below) to investigate how the different guilds of floral visitors interacted along the main axes of reproductive trait variation using qualitative and quantitative measures of plant-pollinator interactions. Specifically, interactions between plants and floral visitors in binary networks were defined as the presence or absence of floral visitation, while in weighted networks, interactions were measured as the the number of visits from each insect species to different plant species. Although floral visitors are not always pollinators and the number of visits does not consider the pollination efficiency of each species [@ballantyne2015], the number of visits does provide valuable information about the relative contributions different floral visitors make to pollination [@vazquez2005; @vazquez2012]. We divided floral visitors into six main taxonomical guilds that differ in life form, behaviour and thus likely play different ecological roles: (i) bees (Hymenoptera-Anthophila), (ii) non-bee-Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Diptera-Syrphidae), (iv) non-syrphid-Diptera (Diptera-non-Syrphidae), (v) Lepidoptera, and (vi) Coleoptera. Moreover, because bees had the greatest number of occurrences in networks  (2,256 records) and the highest frequency of visits, we further broke this group down into the main bee families (Andrenidae, Apidae, Colletidae, Halictidae, and Megachilidae). In addition, we found that *Apis mellifera* was the floral visitor with the largest proportion of occurrences in networks (7.55% of the total). This is consistent with previous research showing that *A. mellifera* was the most frequent floral visitor in a similar dataset of 80 plant-pollinator networks in natural ecosystems [@hung2018]. Hence, to control for the effect of *A. mellifera* on observed visitation patterns of bees, we conducted an analogous analysis with presence-absence of the interaction and number of visits excluding *A. mellifera*. We found that *A. mellifera*, was partly driving some of the observed trends on PC1 (figure S5). However, we did not detect major differences on PC2 and PC3.

We used Bayesian generalised linear mixed models in the R package *brms* [@burkner2017] to model the presence-absence of observed interactions and number of visits as a function of the main axes of plant trait variation and their interactions with floral visitor guilds (e.g., number of visits \~ PC1 x FGs + PC2 x FGs + PC3 x FGs). Because we were interested in possible differences in visitation patterns among floral visitor guilds to plants with different reproductive strategies, we included interactions between the main axes of trait variation (PC1, PC2 and PC3) and the different floral visitor guilds in the model. We added a nested random effect of networks within the study system to capture the variation in networks among studies and within networks. Moreover, we included the phylogenetic covariance matrix as a random factor due to the possible shared evolutionary histories of species and therefore lack of independence among them. We specified for the presence or absence of an interaction and number of visits with Bernoulli and a zero inflated negative binomial distributions, respectively. The models were run with 3,000 iterations with previous 1,000 warm up iterations and with non or very weakly informative priors from the brm function so the priors had negligible influence on the results [@burkner2017]. We set delta ($\Delta$) to 0.99 to avoid divergent transitions and visualized the posterior predictive checks with the function *pp_check* using the *bayesplot* package [@gabry2019].

### (ii) Plant species network roles

We also used Bayesian modelling to investigate how the main axes of trait variation determined different plant species' functional roles within pollination networks. For this, we selected complementary species-level network metrics, commonly applied in bipartite network studies [@dormann2008], with clear ecological interpretations and are relevant for defining species' functional roles in networks. Specifically, the species-level metrics we included were: (i) sum of visits per plant species; (ii) normalised degree, calculated as the number of links per plant species divided by the total possible number of partners; and (iii) specialisation (d') [@bluthgen2006], which measures the deviation of an expected random choice of the available interaction partners and ranges between 0 (maximum generalisation) and 1 (maximum specialisation). Normalised degree and specialisation were calculated with the *specieslevel* function in the R package *bipartite* [@dormann2008]. We modelled each plant species metric (sum of visits, normalised degree and plant specialisation) as a function of the three main axes of trait variation (plant species level metric \~ PC1 + PC2 + PC3). For the different response variables in each model (i.e., each plant species level metric), we used different distribution families (zero inflated negative binomial for the sum of visits, weibull for normalised degree, and zero-one inflated beta for specialisation). Finally, we used the same random factors, model settings and conducted the same posterior predictive checks for each model as detailed above in the 'visitation patterns' section.

Our analyses were conducted on the subset of 60 weighted networks (out of a total of 64 networks) sampled in a unique flowering season and site, which included 556 plant and 1,126 pollinator species. In total, our network dataset (excluding meta- and non-weighted networks) included 2,256 interactions for bees, 1,768 for non-syrphid-dipterans, 845 for syrphids, 437 for lepidopterans, 432 for coleopterans, and 362 for non-bee-hymenopterans. Sampling methods varied across studies but this was accounted for in analyses by including study identity as a  random effect in our models. All analyses were conducted in R version *4.0.3*.

# 3. Results

## (a) Plant strategies analysis

Our phylogenetically informed principal component analysis captured  51.8% and 70.97% of the reproductive trait variation in the first two and three axes, respectively (figure 1 and figure S6). The first principal component (PC1; the flower number - flower size axis) represented 26.72% of the trait variation and indicated a negative correlation between flower number and flower size. The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings \> \|0.5\|; table S3) but style length also contributed moderately (loading = -0.33). One end of this axis comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. For instance, *Cornus florida* occurred at this end of the spectrum, which, on average, has approximately 10,000 flowers, is 7.5 m high, has flowers that are 3 mm wide, styles 3.5 mm long and  two ovules per flower. The other end of this spectrum had species that were short in height, with few large flowers that had long styles and many ovules. For example, *Petunia axillaris* on average, is 0.5 m high and has approximately 10 flowers per plant that are over 50 mm wide with 25 mm long styles and over 200 ovules. The second principal component (PC2; the autonomous selfing - floral display axis) encompassed 25.05% of the trait variation and represented variation from low to high autonomous selfing (i.e., high to low pollinator dependence). The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but all other traits (except ovule number) made moderate contributions (loadings ranging between 0.27 to 0.4; table S3). Generally speaking, species with low autonomous selfing had larger and many flowers with long styles, and were tall. *Zuccagnia punctata* occurred at this extreme of the spectrum, which is a self-incompatible species, completely dependent on animal pollination for seed production, producing  approximately 1,500 flowers per plant with an average height of 3 m and 20 mm long styles. In contrast, species with high autonomous selfing tended to have fewer and smaller flowers with shorter styles, and were shorter in height. *Veronica peregrina* occurred at this end of the spectrum, and is a self-compatible plant that requires little or no pollination by animals, it has approximately 20 flowers per plant, is 0.2 m high and has 0.25 mm long styles. The third principal component (PC3; the style length - autonomous selfing axis) explained a considerable amount of trait variability (19.17%) and represented a positive correlation between style length (loading = -0.66) and autonomous selfing (loading = \|-0.51\|), where species with short styles had low autonomous selfing and species with long styles high autonomous selfing. The remaining traits, apart from ovule number, were all positively correlated with style length and autonomous selfing (loadings from -0.23 to -0.46; table S3). In addition, the pPCA with the subset of species that we had nectar and pollen quantity data for showed that nectar quantity (microlitres of nectar per flower) and pollen grains per flower were positively associated with floral display size. Floral rewards were particularly strongly correlated with flower size but weakly correlated with flower number (figure S4). The first two principal components for this pPCA explained similar variance  (45.52%) and similar associations of traits, despite some variability in the loadings (table S4).

```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, out.width="85%", fig.height=10,fig.align='center',fig.width=10,fig.cap= "\\textbf{Figure 1.} Phylogenetically informed principal component analysis (pPCA) of 1,236 plant species from 28 plant-pollinator network studies. The solid arrows indicate the direction of the different quantitative traits (flower number, plant height, style length, flower size, ovule number and level of autonomous selfing) across the two main axes of trait variation. The length of the arrows indicate the weight of the variables on each principal component and the dashed lines show the opposed direction of trait variation. The icons at both ends of arrows and dashed lines illustrate the extreme form of the trait continuum."}

setwd("~/R_Projects/Reproductive Traits")

#Load libraries
library(viridis) #COLOUR GGPLOT
library(ggplot2)
library(broman) #crayon colours
library(cowplot)
library(magick)

>>>>

\vspace{-1.4cm}

@@

\textbf{Jose B. Lanuza$^{1,2}$} \orcidlink{0000-0002-0287-409X}, \textbf{Romina Rader$^{1}$} \orcidlink{0000-0001-9056-9118}, \textbf{Jamie Stavert$^{3}$} \orcidlink{0000-0002-2103-5320}, \textbf{Liam K. Kendall$^{4}$} \orcidlink{0000-0002-0671-0121}, \textbf{Manu E. Saunders$^{1}$} \orcidlink{0000-0003-0645-8277} \textbf{and Ignasi Bartomeus$^{2}$} \orcidlink{0000-0001-7893-4389}

\small

^1^ School of Environmental and Rural Science, University of New England, Armidale, New South Wales 2350, Australia. ^2^ Estación Biológica de Doñana (EBD-CSIC), E-41092 Seville, Spain. ^3^ Department of Conservation \| Te Papa Atawhai, Auckland, New Zealand. ^4^ Centre for Environmental and Climate Science, Lund University, Sölvegatan 37, S-223 62 Lund, Sweden.

\textbf{Author for correspondence:} Jose B. Lanuza \| [barragansljose\@gmail.com](mailto:barragansljose@gmail.com){.email}

**Acknowledgements.** We thank all researchers that made their data openly available and sent it upon request. We also thank Bryony Wilcox, Greg Bible, Mercedes Sanchez-Lanuza and David Ragel for their help with data collection. We thank Jason Tylianakis for his insightful comments on the manuscript before submission. Finally, we thank Marcos Méndez, Susan Kalisz, Amanda Benoit, May Berenbaum and an anonymous reviewer for thoughtful feedback and useful comments on earlier versions of this manuscript.

**Conflict of interest declaration.** We declare we have no competing interests.

**Authors' contributions.** JBL, RR and IB designed the study. JBL collated the data and conducted analysis with guidance of JS, LKK and IB. JBL wrote the manuscript with contributions of all authors.

**Data accessibility.** All data and code used to conduct this study are available in the online repositories of Zenodo (<https://doi.org/10.5281/zenodo.6705621>) and Github (<https://github.com/JoseBSL/Reproductive-traits>).
